---
title: "GDP School Years Comparison"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

Questions of interest:

Firstly, is the number of years children spend in school associated with a country's wealth and if so to what extent?  How does this relationship vary according to gender and generation of worker?  Secondly, what social factors are associated with a nation's wealth, particularly with regard to the involvement of women in the workforce?

Reasons for choice of data:

The pursuit of wealth on the one hand and good good public services on the other are sometimes portrayed as being at odds with each other. But in many cases the the two go hand in hand. It is therefore of interest to explore the link between education and a nation's wealth as well as other linked social factors. Here we chose to look at gross domestic product per person at purchasing power parity (in units of $1000 US dollars) as it seemed to provide the most comparable figure for how wealthy an average individual citizen would feel in their day to day lives.  We compared this with average number of years spent in school, as a measure of how much a country's government wishes to invest in its citizens.  The relationship is expected to be a circular one in the sense that the wealthier a country is, the more it can afford to spend on education and the more it spends on education, the wealthier one might expect it to become.  

The education data set was also chosen because the gapminder.org website has data going back decades on how many years individuals spent in school, broken down by gender and age, in neat ten year bands. It is of interest to find out which generation's time spent in school is most highly correlated with GDP because it might shed light on how quickly a country wishing to raise the school leaving age might expect a return on its investment, or alternatively, how quickly a growing economy chooses to make such an investment.

The interaction of these two factors with other social factors, particularly those affecting women, are also of interest, because of the potential interactions between education of young girls with their subsequent role as mothers. Decisions about timing of children, participation in the workforce, child mortality and fertility rates etc. are all potentially impacted by how long a girl stays in school and thus have wider societal and economic implications for a country.

We looked at data from 1990 and 2015 because that is the time span covered by the Gapminder data set of this particular GDP measure.  It is also a neat quarter of a century period and the years sit either side of a period of huge economic growth for a lot of the world's emerging economies.

R Packages used for this analysis:

``` {r packages, message = FALSE, warning = FALSE}
library(GGally)
library(skimr)
library(moderndive)
library(olsrr)
library(psych)
library(psychometric)
library(tidyverse)
library(gridExtra)
```

The following sections are the data wrangling done with the raw data as each column was contained in a different file.

We haven't uploaded all the raw data files, just the tidy result, so running this code won't work, but let us know if you'd like to see all the raw files.

``` {r wrangling1}

#  Pull each file and select the data for the year 2015
GDP <- data.frame(country = alternative_gdp_per_capita_ppp_wb$country,
                   gdp = alternative_gdp_per_capita_ppp_wb$"2015")

# select the mean years of 192 countries' men and women spent in schools, recorded in 2015, here Gapminder divided them into 6 groups according to the generation, starting from 15 years old, every 10 years seen as a group
men15_24 <- mean_years_in_school_men_15_to_24_years %>%
  dplyr::select("country", "2015")
men25_34 <- mean_years_in_school_men_25_to_34_years %>%
  dplyr::select("country", "2015")
men35_44 <- mean_years_in_school_men_35_to_44_years %>%
  dplyr::select("country", "2015")
men45_54 <- mean_years_in_school_men_45_to_54_years %>%
  dplyr::select("country", "2015")
men55_64 <- mean_years_in_school_men_55_to_64_years %>%
  dplyr::select("country", "2015")
men65_plus <- mean_years_in_school_men_65_plus_years %>%
  dplyr::select("country", "2015")
women15_24 <- mean_years_in_school_women_15_to_24_years %>%
  dplyr::select("country", "2015")
women25_34 <- mean_years_in_school_women_25_to_34_years %>%
  dplyr::select("country", "2015")
women35_44 <- mean_years_in_school_women_35_to_44_years %>%
  dplyr::select("country", "2015")
women45_54 <- mean_years_in_school_women_45_to_54_years %>%
  dplyr::select("country", "2015")
women55_64 <- mean_years_in_school_women_55_to_64_years %>%
  dplyr::select("country", "2015")
women65_plus <- mean_years_in_school_women_65_plus_years %>%
  dplyr::select("country", "2015")

# Combine all the files and convert the GDP figure into a consistent numerical format

data <- GDP %>%
  mutate(k = str_detect(gdp, "k")) %>%
  mutate(gdp = as.numeric(str_remove(gdp, "k"))) %>%
  mutate(gdp = ifelse(k, gdp, gdp/1000)) %>%
  dplyr::select(-k) %>%
  inner_join(women15_24, by = "country") %>%
  rename(women15_24 = "2015") %>%
  inner_join(women25_34, by = "country") %>%
  rename(women25_34 = "2015") %>%
  inner_join(women35_44, by = "country") %>%
  rename(women35_44 = "2015") %>%
  inner_join(women45_54, by = "country") %>%
  rename(women45_54 = "2015") %>%
  inner_join(women55_64, by = "country") %>%
  rename(women55_64 = "2015") %>%
  inner_join(women65_plus, by = "country") %>%
  rename(women65_plus = "2015") %>%
  inner_join(men15_24, by = "country") %>%
  rename(men15_24 = "2015") %>%
  inner_join(men25_34, by = "country") %>%
  rename(men25_34 = "2015") %>%
  inner_join(men35_44, by = "country") %>%
  rename(men35_44 = "2015") %>%
  inner_join(men45_54, by = "country") %>%
  rename(men45_54 = "2015") %>%
  inner_join(men55_64, by = "country") %>%
  rename(men55_64 = "2015") %>%
  inner_join(men65_plus, by = "country") %>%
  rename(men65_plus = "2015")

# Convert to tidy format

data <- data %>%
  pivot_longer(cols = 2:14, names_to = "variable", values_to = "value")

# Write data to csv file

write.csv(data, "GDP_school_years_2015.csv", row.names = FALSE)

```

``` {r wrangling2}

#  Pull each file and select the data for the year 1990
GDP <- data.frame(country = alternative_gdp_per_capita_ppp_wb$country,
                   gdp = alternative_gdp_per_capita_ppp_wb$"1990")
men15_24 <- mean_years_in_school_men_15_to_24_years %>%
  dplyr::select("country", "1990")
men25_34 <- mean_years_in_school_men_25_to_34_years %>%
  dplyr::select("country", "1990")
men35_44 <- mean_years_in_school_men_35_to_44_years %>%
  dplyr::select("country", "1990")
men45_54 <- mean_years_in_school_men_45_to_54_years %>%
  dplyr::select("country", "1990")
men55_64 <- mean_years_in_school_men_55_to_64_years %>%
  dplyr::select("country", "1990")
men65_plus <- mean_years_in_school_men_65_plus_years %>%
  dplyr::select("country", "1990")
women15_24 <- mean_years_in_school_women_15_to_24_years %>%
  dplyr::select("country", "1990")
women25_34 <- mean_years_in_school_women_25_to_34_years %>%
  dplyr::select("country", "1990")
women35_44 <- mean_years_in_school_women_35_to_44_years %>%
  dplyr::select("country", "1990")
women45_54 <- mean_years_in_school_women_45_to_54_years %>%
  dplyr::select("country", "1990")
women55_64 <- mean_years_in_school_women_55_to_64_years %>%
  dplyr::select("country", "1990")
women65_plus <- mean_years_in_school_women_65_plus_years %>%
  dplyr::select("country", "1990")

# Combine all the files and convert the GDP figure into a consistent numerical format

data <- GDP %>%
  mutate(k = str_detect(gdp, "k")) %>%
  mutate(gdp = as.numeric(str_remove(gdp, "k"))) %>%
  mutate(gdp = ifelse(k, gdp, gdp/1000)) %>%
  dplyr::select(-k) %>%
  inner_join(women15_24, by = "country") %>%
  rename(women15_24 = "1990") %>%
  inner_join(women25_34, by = "country") %>%
  rename(women25_34 = "1990") %>%
  inner_join(women35_44, by = "country") %>%
  rename(women35_44 = "1990") %>%
  inner_join(women45_54, by = "country") %>%
  rename(women45_54 = "1990") %>%
  inner_join(women55_64, by = "country") %>%
  rename(women55_64 = "1990") %>%
  inner_join(women65_plus, by = "country") %>%
  rename(women65_plus = "1990") %>%
  inner_join(men15_24, by = "country") %>%
  rename(men15_24 = "1990") %>%
  inner_join(men25_34, by = "country") %>%
  rename(men25_34 = "1990") %>%
  inner_join(men35_44, by = "country") %>%
  rename(men35_44 = "1990") %>%
  inner_join(men45_54, by = "country") %>%
  rename(men45_54 = "1990") %>%
  inner_join(men55_64, by = "country") %>%
  rename(men55_64 = "1990") %>%
  inner_join(men65_plus, by = "country") %>%
  rename(men65_plus = "1990")

# convert to tidy format

data <- data %>%
  pivot_longer(cols = 2:14, names_to = "variable", values_to = "value")

# Write data to csv file

write.csv(data, "GDP_school_years_1990.csv", row.names = FALSE)

```

Combine 1990 and 2015 data files into one

``` {r combineFiles}
school2015 <- read.csv("GDP_school_years_2015.csv")
school1990 <- read.csv("GDP_school_years_1990.csv")

school2015$year <- 2015
school1990$year <- 1990

school <- rbind(school2015, school1990)

# save as one file

school <- write.csv(school, "GDP_school_years.csv", row.names = FALSE)

```

Prepare data on child mortality, fertility rate, murders, suicides and female participation in the workforce for 1990 and 2015.  More data wrangling. Skip to the next code chunk to run without the original files. Note that this section was actually written after the first phase of analysis, but is moved here for convenience.

``` {r otherData}
# import child mortality data

child_mort <- child_mortality_0_5_year_olds_dying_per_1000_born %>%
  dplyr::select("country", "1990", "2015") %>%
  rename(child_mort_1990 = "1990", child_mort_2015 = "2015")

# import fertility data

fert <- children_per_woman_total_fertility %>%
  dplyr::select("country","1990", "2015") %>%
  rename(fert_1990 = "1990", fert_2015 ="2015")

# import data on female participation in the workforce

fem_part <- females_aged_15plus_labour_force_participation_rate_percent %>%
  dplyr::select("country", "1990", "2015") %>%
  rename(fem_part_1990 = "1990", fem_part_2015 = "2015")

# import data on murders

murders <- murder_per_100000_people %>%
  dplyr::select("country", "1990", "2015") %>%
  rename(murders_1990 = "1990", murders_2015 = "2015")

# import data on suicides

suicides <- suicide_per_100000_people %>%
  dplyr::select("country", "1990", "2015") %>%
  rename(suicides_1990 = "1990", suicides_2015 = "2015")

# take most highly correlated column from the school years 2015 data set

school_fem_2015 <- school2015 %>%
  dplyr::select("country", "gdp" ,"women25_34") %>%
  rename(gdp_2015 = gdp, sch_yrs_2015 = women25_34)

# combine into one file with the same column from the school years 1990 data set

gdp <- school1990 %>%
  dplyr::select("country", "gdp", "women25_34") %>%
  rename(gdp_1990 = gdp, sch_yrs_1990 = women25_34) %>%
  right_join(school_fem_2015, by = "country") %>%
  mutate(log_gdp_1990 = log(gdp_1990), log_gdp_2015 = log(gdp_2015)) %>%
  left_join(child_mort, by ="country") %>%
  left_join(fert, by = "country") %>%
  left_join(fem_part, by = "country") %>%
  left_join(murders, by = "country") %>%
  left_join(suicides, by = "country") %>%
  dplyr::select(country, gdp_1990, gdp_2015, log_gdp_1990, log_gdp_2015, everything())

# convert to tidy format

gdp <- gdp %>%
  pivot_longer(cols = 2:17, names_to = "variable", values_to = "value")

# save to data file

write.csv(gdp, "GDP_1990_2015.csv", row.names = FALSE)

```

Combine all data files into one tidy file with a column to identify the original source file and another column for the year of the data set.

``` {r combineNewData}

gdp <- read.csv("GDP_1990_2015.csv")
school <- read.csv("GDP_school_years.csv")

gdp <- gdp %>%
  mutate(year = str_sub(variable, -4), file = "gdp")
school <- school %>%
  mutate(file = "school")
final <- rbind(school, gdp)

# save to data file

write.csv(final, "Group_20.csv", row.names = FALSE)

```



Analysis starts here.  The rest of the code should run with the Group_20.csv file only.



```{r importData}

data <- read.csv("Group_20.csv")

# separate school data by year and widen data

school2015 <- data %>%
  filter(year == 2015, file == "school") %>%
  dplyr::select(-year, -file) %>%
  pivot_wider(id_cols = "country", names_from = "variable", values_from = "value")
school1990 <- data %>%
  filter(year == 1990, file == "school") %>%
  dplyr::select(-year, -file) %>%
  pivot_wider(id_cols = "country", names_from = "variable", values_from = "value")

```


Initial exploration using ggpairs and skim

``` {r exploration, message = FALSE, warning = FALSE}

ggpairs(school2015[,-1]) # Country variable removed
ggpairs(school1990[,-1])

# The GDP data is very right skewed, so applying a log transformation

school2015 <- school2015 %>%
  mutate(logGDP = log(gdp))
school1990 <- school1990 %>%
  mutate(logGDP = log(gdp))
ggpairs(school2015[,-c(1:2)])
skim(school2015) # get useful summary statistics
ggpairs(school1990[,-c(1:2)])
skim(school1990) # get useful summary statistics

```
Plot boxplots of years in education for different generations

``` {r boxplots}

data %>%
  filter(file == "school", variable != "gdp") %>%
  mutate(Gender = ifelse(str_detect(variable, "women"), "Women", "Men")) %>%
  mutate(ageGroup = str_remove_all(variable, "[women]")) %>%
  ggplot(aes(ageGroup, value, fill = Gender)) +
  geom_boxplot() +
  scale_fill_manual(values = c("orange","limegreen")) +
  theme(axis.text.x = element_text(angle = -45)) +
  facet_wrap(~year, ncol = 2) +
  labs(x = "Age group", y = "Years in school", title = "Fig 1. Years spent in school by generation and gender: 1990 and 2015")


```


Calculate the correlations separately for plotting

``` {r}

# Calculate correlations with logGDP

cor2015 <- data.frame(cor(school2015[,3:14],school2015$logGDP))
names(cor2015) <- "Correlation"
school1990 <- na.omit(school1990)
cor1990 <- data.frame(cor(school1990[,3:14],school1990$logGDP))
names(cor1990) <- "Correlation"

# Calculate confidence intervals for correlations - these assume independence so are not valid for pairwise comparisons

CI2015 <- vapply(cor2015[,1], CIr, numeric(2), n = nrow(school2015))
CI1990 <- vapply(cor1990[,1], CIr, numeric(2), n = nrow(school1990))

```

Commentary on pairs plots:

We are dealing with two different types of variables. For the diagonal, these are density plots of men and women in different age groups (numerical variables). Under diagonal here we have scatter plots, when we look at the first column, they are GDP versus different generational group for both men and women. Also,From the graph we can know the correlation between each age range and GDP for males and females and the correlation between each variable， which are corresponding to the slope of the scatter plots.

After given the log transformation, we pay attention to the last column of the plot. 

In 2015 we find correlation coefficients between 0.684 to 0.788, which suggests a fairly strong  positive relationship between GDP and the number of years children spend in school. For the younger people (age under 45), the correlation is higher than the older people; and there is little difference between the genders.

In 1990,  the correlation coefficients are very similar to 2015. At that time, it was clear from the distributions that older women (age above 45) spent less time in school, where as in men this difference is only really apparent over the age of 65.

We can see from the box plots how the median years spent in school, and the range of the data has increased for all generations over the 25 years between the two data sets for both men and women.

In 2015, logGDP has the largest correlation coefficient of 0.788 with the average number of years of schooling for women aged 25 to 34. The figure for men of 0.783 is also the highest for this age bracket.

In 1990, we also see those aged 25 to 34 most closely correlated to GDP.

To make this clearer we plot a bar chart showing correlations from 1990 data set of each age group with log GDP and scatterplot of highest correlating group

``` {r plot1990}

# Scatterplot of women aged 25 to 34 against logGDP

school1990 %>%
  na.omit() %>% # Remove missing values
  ggplot(aes(women25_34, logGDP)) + geom_point() + geom_smooth(method = "lm", se = FALSE, colour = "limegreen") +
  labs(x = "Years in school - women aged 25-34", title = "Data from 1990") 

# Bar chart of correlations of each generation's years in school with log GDP
cor1990 %>%
  rownames_to_column("Group") %>%
  ggplot(aes(Group, Correlation)) + 
    geom_col(fill = c(rep("limegreen",6),rep("orange",6))) + 
    theme(axis.text.x = element_text(angle = -45)) +
    labs(y = "Correlation with log GDP", title = "Data from 1990") + 
    geom_errorbar(aes(ymax = CI1990[2,], ymin = CI1990[1,]), width = 0.5)
```

The scattergraph shows that a country's log GDP level is positively correlated with the number of years of schooling of women aged 25 to 34. But at the same time there are some outliers, for example the country with the highest GDP in the graph has only a moderate number of years of schooling for women aged 25-34.

The bar graph shows the correlation between log GDP and the number of years of schooling for each generation of the population, which can be seen to rise and then fall as the age of the population increases. The maximum point of correlation occurs between the ages of 25 and 34 for both males and females, while the minimum point occurs after the age of 65.

Note that the error bars show the 95% confidence interval for the correlation between each group and log GDP, but they do not take into account correlation between groups and are therefore not pairwise comparable.

And repeated for 2015

``` {r plot2015}

#Scatterplot of women aged 25 to 34 against logGDP

school2015 %>%
  na.omit() %>%
  ggplot(aes(women25_34, logGDP)) + geom_point() + geom_smooth(method = "lm", se = FALSE, colour = "limegreen") +
  labs(x = "Years in school - women aged 25-34", title = "Data from 2015",y="log (GDP per capita / 1000 USD)")

# Bar chart of correlations of each generation's years in school with log GDP
cor2015 %>%
  rownames_to_column("Group") %>%
  ggplot(aes(Group, Correlation)) + 
  geom_col(fill = c(rep("limegreen",6),rep("orange",6))) + 
  theme(axis.text.x = element_text(angle = -45)) +
  labs(y = "Correlation with log GDP", title = "2015 data") + 
  geom_errorbar(aes(ymax = CI2015[2,], ymin = CI2015[1,]), width = 0.5)
```
These two figures are from 2015. Again, the log GDP and years of schooling for women aged 25 to 34 are shown as a scattergraph, as well as the correlation coefficients between log GDP and years of schooling for each age group and gender of the population in a bar plot.

The results are broadly the same as 1990, although it appears that more data is now concentrated on the right hand side of the graph, indicating that the number of years spent in school has increased for some economies.

``` {r}
# Print all four graphs as one

p1<-school1990 %>%
  na.omit() %>% # Remove missing values
  ggplot(aes(women25_34, logGDP)) + geom_point() + geom_smooth(method = "lm", se = FALSE, colour = "limegreen") + labs(x = "Years in school - women aged 25-34", title = "Data from 1990",y="log (GDP pp / 1000 USD)")

p2<-school1990 %>%
  na.omit() %>% # Remove missing values
  ggplot(aes(men25_34, logGDP)) + geom_point() + geom_smooth(method = "lm", se = FALSE, colour = "limegreen") +
  labs(x = "Years in school - men aged 25-34", title = "Data from 1990",y="log (GDP pp / 1000 USD)")


p3<-school2015 %>%
  na.omit() %>%
  ggplot(aes(women25_34, logGDP)) + geom_point() + geom_smooth(method = "lm", se = FALSE, colour = "limegreen") +
  labs(x = "Years in school - women aged 25-34", title = "Data from 2015",y="log (GDP pp / 1000 USD)")

p4<-school2015 %>%
  na.omit() %>%
  ggplot(aes(men25_34, logGDP)) + geom_point() + geom_smooth(method = "lm", se = FALSE, colour = "limegreen") +
  labs(x = "Years in school - men aged 25-34", title = "Data from 2015",y="log (GDP pp / 1000 USD)")

grid.arrange(p1,p2,p3,p4,nrow=2)

```

We now test the largest and smallest differences in correlation between generations for pairwise significance

``` {r corrSigTests1}

# test for significance between highest and lowest correlations

# biggest difference in 2015

r.test(n = nrow(school2015), r12 = cor2015["women25_34",], r13 = cor2015["women65_plus",], r23 = cor(school2015$women25_34, school2015$women65_plus))
cat("  \n  \n")

# smallest difference in 2015

r.test(n = nrow(school2015), r12 = cor2015["women25_34",], r13 = cor2015["women15_24",], r23 = cor(school2015$women25_34, school2015$women15_24))
cat("  \n  \n")

# biggest difference in 1990

r.test(n = nrow(school1990), r12 = cor1990["women25_34",], r13 = cor1990["women65_plus",], r23 = cor(school1990$women25_34, school1990$women65_plus))
cat("  \n  \n")

# smallest difference in 1990

r.test(n = nrow(school1990), r12 = cor1990["women25_34",], r13 = cor1990["women15_24",], r23 = cor(school1990$women25_34, school1990$women15_24))
cat("  \n  \n")

```

Largest differences (between the 25 to 34 age groups and the over 65s) are highly significant in both 1990 and 2015. The smallest differences between 15 to 24 year olds and 25 to 34 years olds are not significant at 0.05 threshold for either 1990 and 2015.


``` {r corrSigTests2}

# test for significance between highest and lowest correlations

# 25 to 34 year old women and men in 2015

r.test(n = nrow(school2015), r12 = cor2015["women25_34",], r13 = cor2015["men25_34",], r23 = cor(school2015$women25_34, school2015$men25_34))
cat("  \n  \n")

#  65 plus women and men in 1990

r.test(n = nrow(school2015), r12 = cor2015["women65_plus",], r13 = cor2015["men65_plus",], r23 = cor(school2015$women65_plus, school2015$men65_plus))
cat("  \n  \n")

# 25 to 34 year old women and men in 1990

r.test(n = nrow(school1990), r12 = cor1990["women25_34",], r13 = cor1990["men25_34",], r23 = cor(school1990$women25_34, school1990$men25_34))
cat("  \n  \n")

#  65 plus women and men in 1990

r.test(n = nrow(school1990), r12 = cor1990["women65_plus",], r13 = cor1990["men65_plus",], r23 = cor(school1990$women65_plus, school1990$men65_plus))
cat("  \n  \n")

```
We find no significant difference between the correlations for men and women, aged either 25-34 or 65+ in either year.

We now attempt to fit a simple linear regression for highest correlating group.  We do not attempt to fit any multiple linear regressions with this data set because of the high level of multicolinearity between the covariates.

``` {r lm}

# for 2015 data
mod2015f <- lm(logGDP ~ women25_34, data = school2015) # Linear regression of logGDP and women aged 25-34 years
summary(mod2015f) # Get regression results
get_regression_table(mod2015f)
get_regression_summaries(mod2015f)
#

# for 1990 data
mod1990f <- lm(logGDP ~ women25_34, data = school1990) # model
summary(mod1990f) # results
get_regression_table(mod1990f)
get_regression_summaries(mod1990f)

```


$${\widehat{y}} = \widehat{\alpha} + \widehat{\beta}x_{1}$$

where y denotes logGDP and X_1 denotes years of schooling for women aged 25 to 34.

For the 2015 data, alpha hat has a value of 0.0035, while beta hat has a value of 0.2348, suggesting that for every additional year of schooling a women aged 25-34 received, we can expect log GDP per capita (in thousands US dollars) to increase by 0.2348. The R-squared is 0.6363, with a p-value of less than 2.2e-16.The R-square indicates that years of schooling explains 63.63% of log GDP, while a p-value of less than 2.2e-16 indicates that the significance test is passed.

For the 1990 data, alpha hat has a value of 0.0482, while beta hat has a value of 0.2768, while the R-squared is 0.6203, with a p-value of less than 2.2e-16.The R-square indicates that years of schooling explains 62.03% of log GDP, while a p-value of less than 2.2e-16 indicates that the significance test is passed.

The 95% confidence intervals for beta hat for 2015 (0.208, 0.261) and for 1990 (0.242, 0.311) overlap, suggesting that there may not have been a significant change in the parameter between 1990 and 2015.

Check model assumptions for 2015 data

``` {r assumptions}
get_regression_points(mod2015f) %>%
  ggplot(aes(logGDP_hat, residual)) + geom_point() + geom_hline(yintercept = 0, colour = "limegreen")
get_regression_points(mod2015f) %>%
  ggplot(aes(women25_34, residual)) + geom_point() + geom_hline(yintercept = 0, colour = "limegreen")
get_regression_points(mod2015f) %>%
  ggplot(aes(sample = residual)) + stat_qq()
```

Observing the residuals vs fitted values, it is clear that the residuals do appear randomly scatter around zero. Accessing model fit, and the QQ plot shows the dots sitting on a straight line. 

The three plots suggest that the mean of the error is zero, the variance is constant, the variables satisfy a linear relationship, and the assumption of normality is correct.


Recreate the models but for men

``` {r lm1}
mod2015m <- lm(logGDP ~ men25_34, data = school2015)
summary(mod2015m)
get_regression_table(mod2015m)
get_regression_summaries(mod2015m)
exp(get_regression_table(mod2015m)$estimate[2])

mod1990m <- lm(logGDP ~ men25_34, data = school1990)
summary(mod1990m)
get_regression_table(mod1990m)
get_regression_summaries(mod1990m)
exp(get_regression_table(mod1990m)$estimate[2])
```

$${\widehat{y}} = \widehat{\alpha} + \widehat{\beta}x_{2}$$
where y denotes logGDP and X_2 denotes years of schooling for men aged 25 to 34.

For the 2015 data, alpha hat has a value of -0.8954, while beta hat has a value of 0.3125, while the R-squared is 0.6227, with a p-value of less than 2.2e-16.The R-square indicates that years of schooling explains 62.27% of logGDP, while a p-value of less than 2.2e-16 indicates that the significance test is passed.

For the 1990 data, alpha hat has a value of -0.5912, while beta hat has a value of 0.3349, while the R-squared is 0.6124, with a p-value of less than 2.2e-16.The R-square indicates that years of schooling explains 61.24% of logGDP, while a p-value of less than 2.2e-16 indicates that the significance test is passed.

Once again, the beta hat confidence intervals of (0.276, 0.349) for 2015 and (0.293, 0.377) overlap.

Check model assumptions

``` {r assumptions}
get_regression_points(mod2015m) %>%
  ggplot(aes(logGDP_hat, residual)) + geom_point() + geom_hline(yintercept = 0, colour = "orange")
get_regression_points(mod2015m) %>%
  ggplot(aes(men25_34, residual)) + geom_point() + geom_hline(yintercept = 0, colour = "orange")
get_regression_points(mod2015m) %>%
  ggplot(aes(sample = residual)) + stat_qq()
```

These three graphs also show that the model is consistent with the assumptions, although the QQ plot is not quite as straight as for the female model.

Coefficients table for poster:

``` {r}

ModelCoeffs <- data.frame(Year = c(2015, 2015, 1990, 1990),
                          Gender = c("Women", "Men", "Women", "Men"),
                          Coefficient = c(get_regression_table(mod2015f)$estimate[2],
                              get_regression_table(mod2015m)$estimate[2],
                              get_regression_table(mod1990f)$estimate[2],
                              get_regression_table(mod1990m)$estimate[2]),
                          Lower_CI = c(get_regression_table(mod2015f)$lower_ci[2],
                              get_regression_table(mod2015m)$lower_ci[2],
                              get_regression_table(mod1990f)$lower_ci[2],
                              get_regression_table(mod1990m)$lower_ci[2]),
                          Upper_CI = c(get_regression_table(mod2015f)$upper_ci[2],
                              get_regression_table(mod2015m)$upper_ci[2],
                              get_regression_table(mod1990f)$upper_ci[2],
                              get_regression_table(mod1990m)$upper_ci[2]))

ModelCoeffs
  

```

#The coefficients table by gender for 1990 and 2015

Year          | Gender        | Coefficient    | Lower CI     | Upper CI    |
:------------:|:-------------:|:--------------:|:-------------|:------------:
2015          | Women         | 0.235         | 0.208         | 0.261
2015          | Men           | 0.312         | 0.276         | 0.349
1990          | Women         | 0.277         | 0.242         | 0.311
1990          | Men           | 0.335         | 0.293         | 0.377

Table: Model Coefficients by gender for 1990 and 2015

Comparing all the beta estimates we see that an overlap between the confidence intervals of men and women in 1990 is no longer evident in 2015, perhaps indicating a change of relationship worthy of further investigation.

In conclusion we see that there is a highly significant positive association between years spent in school and our measure of GDP. The range of possible values for beta suggest that for every additional year spent in school of the average 25-34 year old, we can expect log GDP per person (in 1000s of US dollars) to increase by between 0.208 and 0.377.  For a country with GDP per capita of $1000 dollars (i.e. log(GDP) = 0), this equates to an increase of between (exp(0.208) - 1) * 1000 = 231 USD and (exp(0.377) - 1) * 1000 = 458 USD per person.

We now turn to some additional covariates:

``` {r importGDPData}
gdp <- read.csv("Group_20.csv")

# widen data
gdp <- gdp %>%
  filter(file == "gdp") %>%
  dplyr::select(-year, -file) %>%
  pivot_wider(id_cols = "country", names_from = "variable", values_from = "value")

```

Explore new data sets

``` {r exploration, message = FALSE, warning = FALSE}

gdp %>%
  skim()
# just look at countries without missing values
gdp %>%
  na.omit() %>%
  skim()
  
gdp %>%
  dplyr::select(-country, -gdp_1990, -gdp_2015) %>%
  ggpairs()

```

Table for poster:

Variable          | log GDP 1990  | log GDP 2015   | 
:----------------:|:-------------:|:--------------:
School Years      |  0.788        |  0.798         |
Child Mortality   | -0.791        | -0.784         |
Fertility         | -0.751        | -0.785         |
Female Work Part  | -0.375        | -0.207         | 

Table: Correlations of social factors with log GDP


As we mentioned above, the interpretation of the matrix is as follows:	The variable names are displayed on the outer edges of the matrix. The boxes along the diagonals display the density plot for each variable. The boxes in the lower left corner display the scatterplot between each variable. The boxes in the upper right corner display the Pearson correlation coefficient between each variable. Here we should pay attention to the first two rows. 

In 2015, there is a strong positive relationship between School years and log GDP, a strong negative relationship between child mortality and log GDP, and fertility rate and log GDP, a very weak negative relationship between female participation in the workforce and log GDP, a weak negative relationship between data on murders and log GDP,a very weak (not significant) positive relationship between suicides and log GDP.

In 1990, there is a strong positive relationship between School years and GDP, a strong negative relationship between child mortality and log GDP, and fertility rate and log GDP, a weak negative relationship between female participation in the workforce and log GDP, a weak negative relationship between data on murders and log GDP,a weak positive (this time significant) relationship between suicides and log GDP.

From these, we could draw the conclusion that the GDP may be most associated with the school years, child mortality, and fertility rate.

First we fit full models with all the covariates and then implement forward stepwise regression based on significance level.  Our intention is to find covariates that may explain more of the variation in the response variable than we have so far found up until now. We then build the multivariate regression models suggested by the stepwise process.

``` {r lm2}

# full models

mod_1990_2 <- lm(log_gdp_1990 ~ sch_yrs_1990 + child_mort_1990 + fert_1990 + fem_part_1990 + murders_1990 + suicides_1990, data = gdp)
summary(mod_1990_2)

mod_2015_2 <- lm(log_gdp_2015 ~ sch_yrs_2015 + child_mort_2015 + fert_2015 + fem_part_2015 + murders_2015 + suicides_2015, data = gdp)
summary(mod_2015_2)

# implement forward stepwise regression based on significance level

ols_step_forward_p(mod_1990_2)
ols_step_forward_p(mod_2015_2)

# rebuild models based on most significant indicators (excluding murders due to high number of missing values)

mod_1990_3 <- lm(log_gdp_1990 ~ child_mort_1990 + sch_yrs_1990 + fem_part_1990, data = gdp)
summary(mod_1990_3)

mod_2015_3 <- lm(log_gdp_2015 ~ child_mort_2015 + sch_yrs_2015 + fert_2015, data = gdp)
summary(mod_2015_3)

```

For the full 1990 model, three variables did not pass the significance test according to the regression results, namely female participation in the workforce, data on murders and data on suicides. So we started with a forward stepwise linear regression.

Child mortality, school years and female participation in the workforce were retained based on which p values were most significant.

The same steps were carried out for the 2015 data and only the child mortality passed the significance test in the full model.The child mortality, school years, murders and fertility variables were retained in a stepwise regression using p-value magnitude as the criterion, but we exclude murders from the final model due to the number of missing values.

Noticeably, fertility rates have taken the place of female participation in the workforce between 1990 and 2015. Also noticeable that female participation in the workplace, as well as child mortality, are negatively correlated. 

Years in school remains the most significant explanatory variable even with the other variables added.

The adjusted R^2 value of both models is 0.72 suggesting that we have explained only an additional tenth of the variation in the data compared to the simple linear regression model. This could be partially explained by the high level of correlation between some of the covariates.  Female participation in the work place is not strongly correlated with the other variables, but there are strong association between child mortality, fertility and years in school which should make us extremely wary of reading too much into these models.

Check model assumptions for model 3: 2015

``` {r assumptions2015}
get_regression_points(mod_2015_3) %>%
  ggplot(aes(log_gdp_2015_hat, residual)) + geom_point() + geom_hline(yintercept = 0)
get_regression_points(mod_2015_3) %>%
  ggplot(aes(sch_yrs_2015, residual)) + geom_point() + geom_hline(yintercept = 0)
get_regression_points(mod_2015_3) %>%
  ggplot(aes(child_mort_2015, residual)) + geom_point() + geom_hline(yintercept = 0)
get_regression_points(mod_2015_3) %>%
  ggplot(aes(fert_2015, residual)) + geom_point() + geom_hline(yintercept = 0)
get_regression_points(mod_2015_3) %>%
  ggplot(aes(sample = residual)) + stat_qq()
```

Check model assumptions for model 3: 1990

``` {r assumptions1990}
get_regression_points(mod_1990_3) %>%
  ggplot(aes(log_gdp_1990_hat, residual)) + geom_point() + geom_hline(yintercept = 0)
get_regression_points(mod_1990_3) %>%
  ggplot(aes(sch_yrs_1990, residual)) + geom_point() + geom_hline(yintercept = 0)
get_regression_points(mod_1990_3) %>%
  ggplot(aes(child_mort_1990, residual)) + geom_point() + geom_hline(yintercept = 0)
get_regression_points(mod_1990_3) %>%
  ggplot(aes(fem_part_1990, residual)) + geom_point() + geom_hline(yintercept = 0)
get_regression_points(mod_1990_3) %>%
  ggplot(aes(sample = residual)) + stat_qq()
```
The graphs show that the model assumptions hold.

print coefficient summaries

``` {r coefficients}

get_regression_table(mod_1990_3)
get_regression_table(mod_2015_3)


```


For completeness we plot each of the covariates used in the models against the response variable

``` {r graphs, warning = F, message = F}

# plot graphs of explanatory variables against response variable

gdp %>%
  ggplot(aes(child_mort_1990, log_gdp_1990)) + geom_point() + geom_smooth(method = "lm", se = F)
gdp %>%
  ggplot(aes(child_mort_2015, log_gdp_2015)) + geom_point() + geom_smooth(method = "lm", se = F)
gdp %>%
  ggplot(aes(sch_yrs_1990, log_gdp_1990)) + geom_point() + geom_smooth(method = "lm", se = F)
gdp %>%
  ggplot(aes(sch_yrs_2015, log_gdp_2015)) + geom_point() + geom_smooth(method = "lm", se = F)
gdp %>%
  ggplot(aes(fem_part_1990, log_gdp_1990)) + geom_point() + geom_smooth(method = "lm", se = F)
gdp %>%
  ggplot(aes(fert_2015, log_gdp_2015)) + geom_point() + geom_smooth(method = "lm", se = F)
  


```
conclusion

In summary, it can be found that in the analysis using P-values, there is a significant relationship between the time children spend in school and national wealth, even accounting for child mortality, fertility rates and female participation in the workplace. The longer children spend in school, the greater the wealth of a country.
