---
title: "GDP School Years Comparison"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

Questions of interest:

Firstly, is the number of years children spend in school associated with a country's wealth and if so to what extent?  How does this relationship vary according to gender and generation of worker?  Secondly, what social factors are associated with a nation's wealth, particularly with regard to the involvement of women in the workforce?

``` {r packages, message = FALSE, warning = FALSE}
library(GGally)
library(skimr)
library(moderndive)
library(olsrr)
library(psych)
library(psychometric)
library(tidyverse)
```

This section is the data wrangling that I did with the raw data as each column was contained in a different file.

I haven't uploaded all the raw files to GitHub, just the final version, so running this code won't work, but let me know if you'd like to see all the raw files.

``` {r wrangling1}

#  Pull each file and select the data for the year 2015
GDP <- data.frame(country = alternative_gdp_per_capita_ppp_wb$country,
                   gdp = alternative_gdp_per_capita_ppp_wb$"2015")

# select the mean years of 192 countries' men spending in schools in 2015, here Gapminder divided them into 6 groups according to the generation, starting from 15 years old, every 10 years seen as a group
men15_24 <- mean_years_in_school_men_15_to_24_years %>%
  dplyr::select("country", "2015")
men25_34 <- mean_years_in_school_men_25_to_34_years %>%
  dplyr::select("country", "2015")
men35_44 <- mean_years_in_school_men_35_to_44_years %>%
  dplyr::select("country", "2015")
men45_54 <- mean_years_in_school_men_45_to_54_years %>%
  dplyr::select("country", "2015")
men55_64 <- mean_years_in_school_men_55_to_64_years %>%
  dplyr::select("country", "2015")
men65_plus <- mean_years_in_school_men_65_plus_years %>%
  dplyr::select("country", "2015")
women15_24 <- mean_years_in_school_women_15_to_24_years %>%
  dplyr::select("country", "2015")
women25_34 <- mean_years_in_school_women_25_to_34_years %>%
  dplyr::select("country", "2015")
women35_44 <- mean_years_in_school_women_35_to_44_years %>%
  dplyr::select("country", "2015")
women45_54 <- mean_years_in_school_women_45_to_54_years %>%
  dplyr::select("country", "2015")
women55_64 <- mean_years_in_school_women_55_to_64_years %>%
  dplyr::select("country", "2015")
women65_plus <- mean_years_in_school_women_65_plus_years %>%
  dplyr::select("country", "2015")

# Combine all the files and convert the GDP figure into a consistent numerical format

data <- GDP %>%
  mutate(k = str_detect(gdp, "k")) %>%
  mutate(gdp = as.numeric(str_remove(gdp, "k"))) %>%
  mutate(gdp = ifelse(k, gdp, gdp/1000)) %>%
  dplyr::select(-k) %>%
  inner_join(women15_24, by = "country") %>%
  rename(women15_24 = "2015") %>%
  inner_join(women25_34, by = "country") %>%
  rename(women25_34 = "2015") %>%
  inner_join(women35_44, by = "country") %>%
  rename(women35_44 = "2015") %>%
  inner_join(women45_54, by = "country") %>%
  rename(women45_54 = "2015") %>%
  inner_join(women55_64, by = "country") %>%
  rename(women55_64 = "2015") %>%
  inner_join(women65_plus, by = "country") %>%
  rename(women65_plus = "2015") %>%
  inner_join(men15_24, by = "country") %>%
  rename(men15_24 = "2015") %>%
  inner_join(men25_34, by = "country") %>%
  rename(men25_34 = "2015") %>%
  inner_join(men35_44, by = "country") %>%
  rename(men35_44 = "2015") %>%
  inner_join(men45_54, by = "country") %>%
  rename(men45_54 = "2015") %>%
  inner_join(men55_64, by = "country") %>%
  rename(men55_64 = "2015") %>%
  inner_join(men65_plus, by = "country") %>%
  rename(men65_plus = "2015")

# COnvert to tidy format

data <- data %>%
  pivot_longer(cols = 2:14, names_to = "variable", values_to = "value")

# Write data to csv file

write.csv(data, "GDP_school_years_2015.csv", row.names = FALSE)

```

``` {r wrangling2}

#  Pull each file and select the data for the year 1990
GDP <- data.frame(country = alternative_gdp_per_capita_ppp_wb$country,
                   gdp = alternative_gdp_per_capita_ppp_wb$"1990")
men15_24 <- mean_years_in_school_men_15_to_24_years %>%
  dplyr::select("country", "1990")
men25_34 <- mean_years_in_school_men_25_to_34_years %>%
  dplyr::select("country", "1990")
men35_44 <- mean_years_in_school_men_35_to_44_years %>%
  dplyr::select("country", "1990")
men45_54 <- mean_years_in_school_men_45_to_54_years %>%
  dplyr::select("country", "1990")
men55_64 <- mean_years_in_school_men_55_to_64_years %>%
  dplyr::select("country", "1990")
men65_plus <- mean_years_in_school_men_65_plus_years %>%
  dplyr::select("country", "1990")
women15_24 <- mean_years_in_school_women_15_to_24_years %>%
  dplyr::select("country", "1990")
women25_34 <- mean_years_in_school_women_25_to_34_years %>%
  dplyr::select("country", "1990")
women35_44 <- mean_years_in_school_women_35_to_44_years %>%
  dplyr::select("country", "1990")
women45_54 <- mean_years_in_school_women_45_to_54_years %>%
  dplyr::select("country", "1990")
women55_64 <- mean_years_in_school_women_55_to_64_years %>%
  dplyr::select("country", "1990")
women65_plus <- mean_years_in_school_women_65_plus_years %>%
  dplyr::select("country", "1990")

# Combine all the files and convert the GDP figure into a consistent numerical format

data <- GDP %>%
  mutate(k = str_detect(gdp, "k")) %>%
  mutate(gdp = as.numeric(str_remove(gdp, "k"))) %>%
  mutate(gdp = ifelse(k, gdp, gdp/1000)) %>%
  dplyr::select(-k) %>%
  inner_join(women15_24, by = "country") %>%
  rename(women15_24 = "1990") %>%
  inner_join(women25_34, by = "country") %>%
  rename(women25_34 = "1990") %>%
  inner_join(women35_44, by = "country") %>%
  rename(women35_44 = "1990") %>%
  inner_join(women45_54, by = "country") %>%
  rename(women45_54 = "1990") %>%
  inner_join(women55_64, by = "country") %>%
  rename(women55_64 = "1990") %>%
  inner_join(women65_plus, by = "country") %>%
  rename(women65_plus = "1990") %>%
  inner_join(men15_24, by = "country") %>%
  rename(men15_24 = "1990") %>%
  inner_join(men25_34, by = "country") %>%
  rename(men25_34 = "1990") %>%
  inner_join(men35_44, by = "country") %>%
  rename(men35_44 = "1990") %>%
  inner_join(men45_54, by = "country") %>%
  rename(men45_54 = "1990") %>%
  inner_join(men55_64, by = "country") %>%
  rename(men55_64 = "1990") %>%
  inner_join(men65_plus, by = "country") %>%
  rename(men65_plus = "1990")

# convert to tidy format

data <- data %>%
  pivot_longer(cols = 2:14, names_to = "variable", values_to = "value")

# Write data to csv file

write.csv(data, "GDP_school_years_1990.csv", row.names = FALSE)

```

Combine 1990 and 2015 data files into one

``` {r combineFiles}
school2015 <- read.csv("GDP_school_years_2015.csv")
school1990 <- read.csv("GDP_school_years_1990.csv")

school2015$year <- 2015
school1990$year <- 1990

school <- rbind(school2015, school1990)

# save as one file

school <- write.csv(school, "GDP_school_years.csv", row.names = FALSE)

```

Prepare data on child mortality, fertility rate, murders, suicides and female participation in the workforce for 1990 and 2015.  More data wrangling. Skip to the next code chunk to run without the original files.

``` {r otherData}
# import child mortality data

child_mort <- child_mortality_0_5_year_olds_dying_per_1000_born %>%
  dplyr::select("country", "1990", "2015") %>%
  rename(child_mort_1990 = "1990", child_mort_2015 = "2015")

# import fertility data

fert <- children_per_woman_total_fertility %>%
  dplyr::select("country","1990", "2015") %>%
  rename(fert_1990 = "1990", fert_2015 ="2015")

# import data on female participation in the workforce

fem_part <- females_aged_15plus_labour_force_participation_rate_percent %>%
  dplyr::select("country", "1990", "2015") %>%
  rename(fem_part_1990 = "1990", fem_part_2015 = "2015")

# import data on murders

murders <- murder_per_100000_people %>%
  dplyr::select("country", "1990", "2015") %>%
  rename(murders_1990 = "1990", murders_2015 = "2015")

# import data on suicides

suicides <- suicide_per_100000_people %>%
  dplyr::select("country", "1990", "2015") %>%
  rename(suicides_1990 = "1990", suicides_2015 = "2015")

# take most highly correlated column from the school years 2015 data set

school_fem_2015 <- school2015 %>%
  dplyr::select("country", "gdp" ,"women25_34") %>%
  rename(gdp_2015 = gdp, sch_yrs_2015 = women25_34)

# combine into one file with the same column from the school years 1990 data set

gdp <- school1990 %>%
  dplyr::select("country", "gdp", "women25_34") %>%
  rename(gdp_1990 = gdp, sch_yrs_1990 = women25_34) %>%
  right_join(school_fem_2015, by = "country") %>%
  mutate(log_gdp_1990 = log(gdp_1990), log_gdp_2015 = log(gdp_2015)) %>%
  left_join(child_mort, by ="country") %>%
  left_join(fert, by = "country") %>%
  left_join(fem_part, by = "country") %>%
  left_join(murders, by = "country") %>%
  left_join(suicides, by = "country") %>%
  dplyr::select(country, gdp_1990, gdp_2015, log_gdp_1990, log_gdp_2015, everything())

# convert to tidy format

gdp <- gdp %>%
  pivot_longer(cols = 2:17, names_to = "variable", values_to = "value")

# save to data file

write.csv(gdp, "GDP_1990_2015.csv", row.names = FALSE)

```

Combine all data files into one tidy file

``` {r combineNewData}

gdp <- read.csv("GDP_1990_2015.csv")
school <- read.csv("GDP_school_years.csv")

gdp <- gdp %>%
  mutate(year = str_sub(variable, -4), file = "gdp")
school <- school %>%
  mutate(file = "school")
final <- rbind(school, gdp)

# save to data file

write.csv(final, "Group_20.csv", row.names = FALSE)

```



Analysis starts here.  The rest of the code should run with the Group_20.csv file only.



```{r importData}

data <- read.csv("Group_20.csv")

# separate school data by year and widen data

school2015 <- data %>%
  filter(year == 2015, file == "school") %>%
  dplyr::select(-year, -file) %>%
  pivot_wider(id_cols = "country", names_from = "variable", values_from = "value")
school1990 <- data %>%
  filter(year == 1990, file == "school") %>%
  dplyr::select(-year, -file) %>%
  pivot_wider(id_cols = "country", names_from = "variable", values_from = "value")

```


Initial exploration using ggpairs and skim

``` {r exploration, message = FALSE, warning = FALSE}

ggpairs(school2015[,-1]) # Country variable removed
ggpairs(school1990[,-1])

# The GDP data is very right skewed, so applying a log transformation

school2015 <- school2015 %>%
  mutate(logGDP = log(gdp))
school1990 <- school1990 %>%
  mutate(logGDP = log(gdp))
ggpairs(school2015[,-c(1:2)])
skim(school2015) # get useful summary statistics
ggpairs(school1990[,-c(1:2)])
skim(school1990) # get useful summary statistics

```

Calculate the correlations separately for plotting

``` {r}

# Calculate correlations with logGDP

cor2015 <- data.frame(cor(school2015[,3:14],school2015$logGDP))
names(cor2015) <- "Correlation"
school1990 <- na.omit(school1990)
cor1990 <- data.frame(cor(school1990[,3:14],school1990$logGDP))
names(cor1990) <- "Correlation"

# Calculate confidence intervals for correlations - these assume independence so are not valid for pairwise comparisons

CI2015 <- vapply(cor2015[,1], CIr, numeric(2), n = nrow(school2015))
CI1990 <- vapply(cor1990[,1], CIr, numeric(2), n = nrow(school1990))

```

Commentary on pairs plots:

We are dealing with two different types of variables. For the diagonal , these are density plots of men and women in different age groups (numerical variables). Under diagonal here we have scatter plots, when we look at the first column, they are GDP versus different generational group for both men and women. Also,From the graph we can know the correlation between each age range and GDP for males and females and the correlation between each variable， which are corresponding to the slope of the scatter plots.

In 2015, we find correlation coefficients between 0.5 to 0.65, which suggests a moderate  positive relationship between GDP and the number of years children spend in school. For the younger people (age under 45), the correlation relationship is much stronger than the older people; and there is little difference between the gender.

In 1990,  the correlation coefficients (0.50 to 0.59) are smaller than that in 2015. At that time, it was obvious that older women (age above 45) spent less time in school. The richer the country was, the longer time the men in that country could stay at school. Also, there is little difference between different generational men groups.

After given the log transformation, we pay attention to the last column of the plot. 

Here we could draw the same conclusion that, in 2015, logGDP still has the largest correlation coefficient of 0.7976 with the average number of years of schooling for women aged 25 to 34. And the relationship between GDP and the number of years children spend in school had little difference between the gender. But, for younger generation, the correlation is much stronger.

It was a common sense that for both men and women, younger generation spent more years in school.

In 1990,The same thing was younger people had more time in school, the different situation was  people spent less time in school than people in 2015. For women, they had less opportunity to go to school than men. 

The data show that in 1990 logGDP had the highest correlation coefficient of 0.7876 with the average number of years of schooling of women aged 25 to 34.

Bar chart showing correlations from 1990 data set of each age group with log GDP and scatterplot of highest correlating group

``` {r plot1990}

# Scatterplot of women aged 25 to 34 against logGDP

school1990 %>%
  na.omit() %>% # Remove missing values
  ggplot(aes(women25_34, logGDP)) + geom_point() + geom_smooth(method = "lm", se = FALSE, colour = "limegreen") +
  labs(x = "Years in school - women aged 25-34", title = "Data from 1990") 

# Bar chart of correlations of each generation's years in school with log GDP
cor1990 %>%
  rownames_to_column("Group") %>%
  ggplot(aes(Group, Correlation)) + 
    geom_col(fill = c(rep("limegreen",6),rep("orange",6))) + 
    theme(axis.text.x = element_text(angle = -45)) +
    labs(y = "Correlation with log GDP", title = "Data from 1990") + 
    geom_errorbar(aes(ymax = CI1990[2,], ymin = CI1990[1,]), width = 0.5)
```

The scattergraph shows that a country's log GDP level is positively correlated with the number of years of schooling of women aged 25 to 34. But at the same time there are some outliers, for example the country with the highest GDP in the graph have only a moderate number of years of schooling for women aged 25-34.

The bar graph shows the correlation between log GDP and the number of years of schooling for each generation of the population, which can be seen to rise and then fall as the age of the population increases. The maximum point of correlation occurs between the ages of 25 and 34 for both males and females, while the minimum point occurs after the age of 65.

Bar plots here were used to visualize the distributions of categorical variables. By comparing  different generation groups, we could directly draw the conclusion mentioned above. For both men and women, people aged 25 to 34 had largest correlation coefficient.

And repeated for 2015

``` {r plot2015}

#Scatterplot of women aged 25 to 34 against logGDP

school2015 %>%
  na.omit() %>%
  ggplot(aes(women25_34, logGDP)) + geom_point() + geom_smooth(method = "lm", se = FALSE, colour = "limegreen") +
  labs(x = "Years in school - women aged 25-34", title = "2015 data")

# Bar chart of correlations of each generation's years in school with log GDP
cor2015 %>%
  rownames_to_column("Group") %>%
  ggplot(aes(Group, Correlation)) + 
  geom_col(fill = c(rep("limegreen",6),rep("orange",6))) + 
  theme(axis.text.x = element_text(angle = -45)) +
  labs(y = "Correlation with log GDP", title = "2015 data") + 
  geom_errorbar(aes(ymax = CI2015[2,], ymin = CI2015[1,]), width = 0.5)
```
These two figures are from 2015. Again, the log GDP and years of schooling for women aged 25 to 34 are shown as a scattergraph, as well as the correlation coefficients between log GDP and years of schooling for each age group and gender of the population in a bar plot.

The results are broadly the same as 1990, although it appears that more data is now concentrated on the right hand side of the graph, indicating that the number of years spent in school has increased for some economies.

We now test the largest and smallest differences in correlation between generations for pairwise significance

``` {r corrSigTests}

# test for significance between highest and lowest correlations

# biggest difference in 2015

r.test(n = nrow(school2015), r12 = cor2015["women25_34",], r13 = cor2015["women65_plus",], r23 = cor(school2015$women25_34, school2015$women65_plus))
cat("  \n  \n")

# smallest difference in 2015

r.test(n = nrow(school2015), r12 = cor2015["women25_34",], r13 = cor2015["women15_24",], r23 = cor(school2015$women25_34, school2015$women15_24))
cat("  \n  \n")

# biggest difference in 1990

r.test(n = nrow(school1990), r12 = cor1990["women25_34",], r13 = cor1990["women65_plus",], r23 = cor(school1990$women25_34, school1990$women65_plus))
cat("  \n  \n")

# smallest difference in 1990

r.test(n = nrow(school1990), r12 = cor1990["women25_34",], r13 = cor1990["women15_24",], r23 = cor(school1990$women25_34, school1990$women15_24))
cat("  \n  \n")

```

Largest differences are highly significant, smallest differences are not significant at 0.05 threshold




We now attempt to fit a linear model fitted for highest correlating group 

``` {r lm}

# for 2015 data
mod2015f <- lm(logGDP ~ women25_34, data = school2015) # Linear regression of logGDP and women aged 25-34 years
summary(mod2015f) # Get regression results
get_regression_table(mod2015f)
get_regression_summaries(mod2015f)
#

# for 1990 data
mod1990f <- lm(logGDP ~ women25_34, data = school1990) # model
summary(mod1990f) # results
get_regression_table(mod1990f)
get_regression_summaries(mod1990f)

```


$${\widehat{y}} = \widehat{\alpha} + \widehat{\beta}x_{1}$$

where y denotes logGDP and X1 denotes years of schooling for women aged 25 to 34.

For the 2015 data, alpha hat has a value of 0.0035, while beta hat has a value of 0.2348, while the R-squared is 0.6363, with a p-value of less than 2.2e-16.The R-square indicates that years of schooling explains 63.63% of logGDP, while a p-value of less than 2.2e-16 indicates that the significance test is passed.

For the 1990 data, alpha hat has a value of 0.0482, while beta hat has a value of 0.0.2768, while the R-squared is 0.6203, with a p-value of less than 2.2e-16.The R-square indicates that years of schooling explains 62.03% of logGDP, while a p-value of less than 2.2e-16 indicates that the significance test is passed.

Check model assumptions for 2015 data

``` {r assumptions}
get_regression_points(mod2015f) %>%
  ggplot(aes(logGDP_hat, residual)) + geom_point() + geom_hline(yintercept = 0, colour = "limegreen")
get_regression_points(mod2015f) %>%
  ggplot(aes(women25_34, residual)) + geom_point() + geom_hline(yintercept = 0, colour = "limegreen")
get_regression_points(mod2015f) %>%
  ggplot(aes(sample = residual)) + stat_qq()
```

Observing the residuals vs fitted values, it is clear that the residuals do appear randomly scatter around zero. Accessing model fit, and The QQ plot shows the dots sitting on a straight line. 

The three plots each show that the mean of the error is zero, the variance is constant, the variables satisfy a linear relationship, and the assumption of normality is correct.


Recreate the models but for men

``` {r lm1}
mod2015m <- lm(logGDP ~ men25_34, data = school2015)
summary(mod2015m)
get_regression_table(mod2015m)
get_regression_summaries(mod2015m)
exp(get_regression_table(mod2015m)$estimate[2])

mod1990m <- lm(logGDP ~ men25_34, data = school1990)
summary(mod1990m)
get_regression_table(mod1990m)
get_regression_summaries(mod1990m)
exp(get_regression_table(mod1990m)$estimate[2])
```

$${\widehat{y}} = \widehat{\alpha} + \widehat{\beta}x_{2}$$
where y denotes logGDP and X2 denotes years of schooling for men aged 25 to 34.

For the 2015 data, alpha hat has a value of -0.8954, while beta hat has a value of 0.3125, while the R-squared is 0.6227, with a p-value of less than 2.2e-16.The R-square indicates that years of schooling explains 62.27% of logGDP, while a p-value of less than 2.2e-16 indicates that the significance test is passed.

For the 1990 data, alpha hat has a value of -0.5912, while beta hat has a value of 0.3349, while the R-squared is 0.7386, with a p-value of less than 2.2e-16.The R-square indicates that years of schooling explains 73.86% of logGDP, while a p-value of less than 2.2e-16 indicates that the significance test is passed.

Because of the log transformation, here we need to use exp() function to get the real result.


Check model assumptions

``` {r assumptions}
get_regression_points(mod2015m) %>%
  ggplot(aes(logGDP_hat, residual)) + geom_point() + geom_hline(yintercept = 0, colour = "orange")
get_regression_points(mod2015m) %>%
  ggplot(aes(men25_34, residual)) + geom_point() + geom_hline(yintercept = 0, colour = "orange")
get_regression_points(mod2015m) %>%
  ggplot(aes(sample = residual)) + stat_qq()
```

These three graphs also show that the model is also consistent with the assumptions, although the QQ plot is not quite as straight as for the female model.

Conclusion

As we known, schooling tends to be largely financed with public resources across the globe. Through the investigation, there is a moderate  positive relationship between log GDP and the number of years children spend in school, both in 1990 and in 2015. That is the longer time children spent in school, the larger the GDP was, which also suggested the richer the country was. 


``` {r importGDPData}
gdp <- read.csv("Group_20.csv")

# widen data
gdp <- gdp %>%
  filter(file == "gdp") %>%
  dplyr::select(-year, -file) %>%
  pivot_wider(id_cols = "country", names_from = "variable", values_from = "value")

```

Explore new data sets

``` {r exploration, message = FALSE, warning = FALSE}

gdp %>%
  skim()
# just look at countries without missing values
gdp %>%
  na.omit() %>%
  skim()
  
gdp %>%
  dplyr::select(-country, -gdp_1990, -gdp_2015) %>%
  ggpairs()

```

As we mentioned above, the interpretation of the matrix is as follows:	The variable names are displayed on the outer edges of the matrix. The boxes along the diagonals display the density plot for each variable. The boxes in the lower left corner display the scatterplot between each variable. The boxes in the upper right corner display the Pearson correlation coefficient between each variable. Here we should pay attention to the first two rows. In 2015, there is a strong positive relationship between School years and log GDP, a strong negative relationship between child mortality and log GDP, the same as fertility data and log GDP, a vary weak negative relationship between female participation in the workforce and log GDP, a weak negative relationship between data on murders and log GDP,a very weak positive relationship between suicides and log GDP.

In 1990, there is a strong positive relationship between School years and GDP, a strong negative relationship between child mortality and log GDP, the same as fertility data and log GDP, a weak negative relationship between female participation in the workforce and log GDP, a weak negative relationship between data on murders and log GDP,a weak positive relationship between suicides and log GDP.

From these, we could draw the conclusion that the GDP may be associated with the school years, child mortality, and fertility of the women.

Fit new linear models

``` {r lm2}

# full models

mod_1990_2 <- lm(log_gdp_1990 ~ sch_yrs_1990 + child_mort_1990 + fert_1990 + fem_part_1990 + murders_1990 + suicides_1990, data = gdp)
summary(mod_1990_2)

mod_2015_2 <- lm(log_gdp_2015 ~ sch_yrs_2015 + child_mort_2015 + fert_2015 + fem_part_2015 + murders_2015 + suicides_2015, data = gdp)
summary(mod_2015_2)

# implement forward stepwise regression based on significance level

ols_step_forward_p(mod_1990_2)
ols_step_forward_p(mod_2015_2)

# rebuild models based on most significant indicators (excluding murders due to high number of missing values)

mod_1990_3 <- lm(log_gdp_1990 ~ child_mort_1990 + sch_yrs_1990 + fem_part_1990, data = gdp)
summary(mod_1990_3)

mod_2015_3 <- lm(log_gdp_2015 ~ child_mort_2015 + sch_yrs_2015 + fert_2015, data = gdp)
summary(mod_2015_3)

```

For the full 1990 model, three variables did not pass the significance test according to the regression results, namely female participation in the workforce, data on murders and data on suicides. So we started with a forward stepwise linear regression.

Child mortality, school years and female participation in the workforce were retained based on which p values were most significant.

The same steps were carried out for the 2015 data and only the child mortality passed the significance test in the full model.The child mortality, school years, murders and fertility variables were retained in a stepwise regression using p-value magnitude as the criterion, but we exclude murders from the final model due to the number of missing values.

Noticeable that fertility rates have taken the place of female participation in the workforce between 1990 and 2015. Also noticeable that female participation in the workplace is negatively correlated. 

Check model assumptions for model 3: 2015

``` {r assumptions2015}
get_regression_points(mod_2015_3) %>%
  ggplot(aes(log_gdp_2015_hat, residual)) + geom_point() + geom_hline(yintercept = 0)
get_regression_points(mod_2015_3) %>%
  ggplot(aes(sch_yrs_2015, residual)) + geom_point() + geom_hline(yintercept = 0)
get_regression_points(mod_2015_3) %>%
  ggplot(aes(child_mort_2015, residual)) + geom_point() + geom_hline(yintercept = 0)
get_regression_points(mod_2015_3) %>%
  ggplot(aes(fert_2015, residual)) + geom_point() + geom_hline(yintercept = 0)
get_regression_points(mod_2015_3) %>%
  ggplot(aes(sample = residual)) + stat_qq()
```

Check model assumptions for model 3: 1990

``` {r assumptions1990}
get_regression_points(mod_1990_3) %>%
  ggplot(aes(log_gdp_1990_hat, residual)) + geom_point() + geom_hline(yintercept = 0)
get_regression_points(mod_1990_3) %>%
  ggplot(aes(sch_yrs_1990, residual)) + geom_point() + geom_hline(yintercept = 0)
get_regression_points(mod_1990_3) %>%
  ggplot(aes(child_mort_1990, residual)) + geom_point() + geom_hline(yintercept = 0)
get_regression_points(mod_1990_3) %>%
  ggplot(aes(fem_part_1990, residual)) + geom_point() + geom_hline(yintercept = 0)
get_regression_points(mod_1990_3) %>%
  ggplot(aes(sample = residual)) + stat_qq()
```

print coefficient summaries

``` {r coefficients}

get_regression_table(mod_1990_3)
get_regression_table(mod_2015_3)


```



The graphs show that the model assumptions hold.

conclusion

In summary, it can be found that in the analysis using P-values, there is a significant relationship between the time children spend in school and child mortality and national wealth. The longer children spend in school and the lower the child mortality rate, the greater the wealth of a country.

For the variable of time children spend in school, according to the regression results, all else being equal, each year increase in the amount of time children spend in school increases logGDP by 0.088259.

For child mortality, each decrease of 1 increases logGDP by 0.01878



``` {r graphs, warning = F, message = F}

# plot graphs of explanatory variables against response variable

gdp %>%
  ggplot(aes(child_mort_1990, log_gdp_1990)) + geom_point() + geom_smooth(method = "lm", se = F)
gdp %>%
  ggplot(aes(child_mort_2015, log_gdp_2015)) + geom_point() + geom_smooth(method = "lm", se = F)
gdp %>%
  ggplot(aes(sch_yrs_1990, log_gdp_1990)) + geom_point() + geom_smooth(method = "lm", se = F)
gdp %>%
  ggplot(aes(sch_yrs_2015, log_gdp_2015)) + geom_point() + geom_smooth(method = "lm", se = F)
gdp %>%
  ggplot(aes(fem_part_1990, log_gdp_1990)) + geom_point() + geom_smooth(method = "lm", se = F)
gdp %>%
  ggplot(aes(fert_2015, log_gdp_2015)) + geom_point() + geom_smooth(method = "lm", se = F)
  


```
